---
title: "Prioritizing Aquaculture Potential Along the U.S. West Coast"
description: "Using environmental suitability modeling to identify the best Exclusive Economic Zones (EEZs) for developing marine aquaculture."
author: "Ixel"
date: "12/12/2025"
categories: [R, Geospatial Analysis, Aquaculture]
draft: false
format: 
  html:
    embed-resources: true
    standalone: true
    toc: true
    code-fold: true
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---

![**Image:[Commercial Fishing Boats, Santa Barbara Harbor, California, USA.](https://santabarbaraca.gov/things-do/waterfront)).**](figs/HarborfromWflyingW300.png.webp)


Marine aquaculture is rapidly expanding as coastal nations seek sustainable ways to meet seafood demand. Identifying *where* aquaculture should occur is a major challenge optimal sites must balance ecological suitability, production needs, and regulatory boundaries.  


## Research Question 
What Exclusive Economic Zones (eez) on the west Coast of the US are best suited to developing marine aquaculture for several species of oysters and Dungeness crab (*Metacarcinus magister* (Dana, 1852)).


### Data Sources

I combined three geospatial datasets:

- **Bathymetry:** GEBCO 2022 global depth raster  
- **Sea Surface Temperature (SST):** 2008–2012 annual averages  
- **Exclusive Economic Zones:** Cleaned U.S. West Coast EEZ shapefile  

Environmental requirements were taken from fisheries and species-specific literature, including SeaLifeBase [(2025)].

All geospatial processing was performed in **R** using `terra`, `stars`, and `sf`.


### Methods

#### 1. Prepare SST and Depth
I stacked five annual SST rasters (2008–2012), calculated the mean, and converted Kelvin to Celsius:
```{r,include = FALSE}
#| output: false

# Load in Libraries
library(here)
library(testthat)
library(tidyverse)
library(sf)
library(kableExtra)
library(sp)
library(stars) 
library(tmap)
library(terra)
library(viridisLite)
```

```{r,include = FALSE}
#| output: false
# Prepare Data

# Bathymetry
depth <- read_stars(here::here("posts/2025-12-12-EDS223_Blog/aqua_data/depth.tif"))

# Exclusive Economic Zones (eez)
eez <- st_read(here::here("posts/2025-12-12-EDS223_Blog/aqua_data/wc_regions_clean.shp"))

# Sea Surface Temperature (SST) load as a list of .tiff files 
sst <- list.files(path= here::here("posts/2025-12-12-EDS223_Blog/aqua_data"),
                  pattern= "average_annual", # file name matching pattern
                   full.names = TRUE) # Reference entire file name matching this pattern

# Stack all rasters files in folder using rast()
sst <- terra::rast(sst)
names(sst) #checking list
```

```{r,include = FALSE}

# Check and align CRS for multiple datasets
if (st_crs(depth) == st_crs(eez)) {
  print("Coordinate reference system match!")
  
} else{
  warning("CRS mismatch between depth and eez")
  # tranfgotm data to match
  depth <- st_transform(depth, st_crs(eez))

  }
   

if (st_crs(eez)$wkt != crs(sst)) {
  warning("CRS mismatch between eez and sst")
  sst <- project(sst, st_crs(eez)$wkt)  
}
```

```{r}
# Calculate the SST average 
mean_sst <- app(sst[[c("average_annual_sst_2008",
                           "average_annual_sst_2009", 
                           "average_annual_sst_2010",
                           "average_annual_sst_2011", 
                           "average_annual_sst_2012")]], 
                    fun = mean, na.rm = TRUE)

# Convert average SST from Kelvin to Celsius
sst_mean_celsius <- mean_sst - 273.15
```

Bathymetry was cropped and resampled to match SST resolution:

```{r}
# First convert depth stars to SpatRaster
depth_rast <- rast(depth)

depth_resample <- depth_rast %>% 
  crop(sst_mean_celsius) %>% # crop depth to match SST  extent
# Resample to match SST resolution, use method near  
  resample(sst_mean_celsius,
           method = "near") 
```

```{r, include = FALSE}
#| output: true

#Check resolution match 
if (all(dim(depth_resample) == dim(sst_mean_celsius)) &&
    isTRUE(all.equal(res(depth_resample), res(sst_mean_celsius))) &&
    crs(depth_resample) == crs(sst_mean_celsius) &&
    ext(depth_resample) == ext(sst_mean_celsius)) {
  cat("✓ All raster properties match!\n")
} else {
  cat("✗ Some properties do not match\n")
}

# Match depth and SST CRS's
crs(depth_resample) <- crs(sst_mean_celsius)

# Stack
ocean_stack <- c(sst_mean_celsius, depth_resample)
names(ocean_stack) <- c("sst", "depth")
cat("✓ Successfully stacked!\n")
```

## Identify Suitable Habitat for Oysters

**Environmental ranges used:**

  - **Sea Surface Temperature Range:** 11–30°C

  - **Depth Range:** 0–70 m

```{r}
# Reclassify SST (11-30°C suitable for oysters)
sst_suitable <- app(sst_mean_celsius, fun = function(x){
  ifelse(x >= 11 & x <= 30,1,0) # if temperate is between 11- 30 °C, return 1, else 0
})

# Reclassify Suitable depth range (0-70 meters for oysters)
depth_suitable <- app(depth_resample, fun = function(x){
  ifelse(x >= -70 & x <= 0,1,0) # select depths from 0 to 70 meters below surface.
})

# Multiply the two suitability raster

suitable_locations <- sst_suitable * depth_suitable

```

## Quantify Suitable Area Within Each EEZ

I rasterized EEZ polygons, masked suitability value, and computed cell-area:
```{r}
# Rasterize the EEZ data (creates zone IDs)
eez_raster <- rasterize(eez, suitable_locations, field = "rgn")
#Calculate cell area with cellSize()
cell_area <- cellSize(suitable_locations, unit = "km")
# Mask suitable locations to EEZ boundaries
suitable_area <- mask(suitable_locations, eez) * cell_area
# Sum suitable are by EEZ zonal statistics 
eez_suitable <- zonal(suitable_area, eez_raster, fun = "sum", na.rm = TRUE)
```

## Results
**Oyster Suitability**

The highest-ranked region was:

Central California (CA-C)
~4,940 km² of suitable area
(~34% of all suitable habitat)
```{r,  echo = FALSE}

colnames(eez_suitable) <- c("eez_id", "suitable_area_km2")

# join with eez data
eez$suitable_area_km2 <- 
  eez_suitable$suitable_area_km2[match(eez$rgn, 
                                        eez_suitable$eez_id)]

# Rank EEZ data
eez_ranked <- eez[order(-eez$suitable_area_km2), ]

summary_table <- data.frame(
  Rank = 1:nrow(eez_ranked),
  EEZ_Name = eez_ranked$rgn_key,
  Suitable_Area_km2 = round(eez_ranked$suitable_area_km2, 2),
  Percent_of_Total = round(100 * eez_ranked$suitable_area_km2 / 
                           sum(eez_ranked$suitable_area_km2), 1)
)


eez_table <- kable(summary_table,
      caption = "EEZ Rankings by Suitable Aquaculture Area",
      col.names = c("Rank", "EEZ Region", "Suitable Area (km²)", "% of Total"),
      align = c("c", "l", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hold_position"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#4CAF50", color = "white") %>%
  row_spec(1, bold = TRUE, background = "#E8F5E9")  # Highlight top rank

eez_table
```

**Figure 1.** Oyster Aquaculture Suitability by EEZ
```{r, message=FALSE}
#| fig.width: 10
#| fig.height: 8

# Map with tmap basic
tm_basemap("Esri.OceanBasemap", alpha = 0.7) +
tm_shape(eez_ranked) +
  tm_polygons("suitable_area_km2",
              palette = "-mako",
              title = "Suitable Area (km²)",
              border.col = "black",
              border.lwd = 0.5) +
  tm_layout(main.title = "Suitable Oyster Aquaculuture Area by EEZ",
            main.title.size = 1.2,
            legend.outsides = TRUE,
            frame = FALSE) + 
    tm_text("rgn", # Label by region
          size = .8, # Adjust size
          col = "white", # Adjust text color
          fontface = "bold", # Labels are bolded
          xmod = -.5) + # Adjust .5 from the left
  tm_compass(type = "4star", position = c("right", "top")) +
  tm_scalebar(position = c("left", "bottom")) 
  
 
```

# Repeat previous steps with species of choice by creating a Function
![**Image:[© Neil McDaniel](https://www.inaturalist.org/photos/446850127)).**](figs/medium.jpd)

Here we use the Dungeness Crab one of the most important seafood industries along the west coast of North America. Dungeness typical range extends from Alaska's Aleutian Islands to Point Conception, near Santa Barbara, California. Commonly found on soft bottom in very low intertidal areas to a depth of 360 meter and prefer a Sea surface temperature of 9°C

Dungeness Crab Suitable Conditions:

  - **Seas Surface Temperature Range:** 3-19°C

  - **Depth Range:** 0 –360 meters

A general function automated the workflow for any species.

```{r, include = FALSE}

find_suitable_aquaculture <- function(sst_raster, 
                                      depth_raster, 
                                      eez_sf, 
                                      sst_min, 
                                      sst_max, 
                                      depth_min, 
                                      depth_max,
                                      species_name) {
  
# Reclassify SST
sst_suitable <- app(sst_raster, fun = function(x) {
    ifelse(x >= sst_min & x <= sst_max, 1, 0)
  })
  
# Reclassify depth
depth_suitable <- app(depth_raster, fun = function(x) {
    ifelse(x >= depth_min & x <= depth_max, 1, 0)
  })
  
# Find suitable locations (both conditions)
suitable_locations <- sst_suitable * depth_suitable
names(suitable_locations) <- species_name 

# Mask to EEZ
suitable_in_eez <- mask(suitable_locations, eez_sf)
  
# Rasterize EEZ
eez_rasterize <- rasterize(eez_sf, suitable_locations, field = "rgn")
  
# Calculate cell areas
cell_area_km2 <- cellSize(suitable_locations, unit = "km")
 

# Calculate suitable area per cell (commented out for now)
suitable_area <- suitable_in_eez * cell_area_km2
  

# Sum by EEZ zone
eez_suitable_area <- zonal(
  suitable_area, 
  eez_rasterize, 
  fun = "sum", 
  na.rm = TRUE)

colnames(eez_suitable_area) <- c("EEZ_ID", "Suitable_Area_km2")
  
# Join with EEZ data and rank
eez_sf$suitable_area_km2 <-
  eez_suitable_area$Suitable_Area_km2[
    match(eez_sf$rgn, eez_suitable_area$EEZ_ID)]
  eez_ranked <- eez_sf[order(-eez_sf$suitable_area_km2), ]
  
# Create summary table
summary_table <- data.frame(
   Rank = 1:nrow(eez_ranked),
    EEZ = eez_ranked$rgn_key,
    Suitable_Area_km2 = round(eez_ranked$suitable_area_km2, 2)
  )
  
# Return results as a list
results <- list(
   species = species_name,
  sst_suitable = sst_suitable,
   depth_suitable = depth_suitable,
   suitable_locations = suitable_locations,
   suitable_in_eez = suitable_in_eez,
   eez_ranked = eez_ranked,
    summary_table = summary_table
  )
  
 return(results)
}
```

## Dungeness Crab Suitability

Using the generated species-specific function:

```{r}
# Mussels (different requirements)
mussel_results <- find_suitable_aquaculture(
  sst_raster = sst_mean_celsius,
  depth_raster = depth_resample,
  eez_sf = eez,
  sst_min = 3,
  sst_max = 19,
  depth_min = 0,
  depth_max = 360,
  species_name = "Dungeness Crab"
)

# Access results
kable(mussel_results$summary_table)
```

**Figure 2.** Datemussel Suitability by EEZ
```{r, message=FALSE}
#| fig.width: 10
#| fig.height: 8

# Plot Mussel EEZ
tm_basemap("Esri.OceanBasemap", alpha = 0.7) +
  tm_raster() +
  tm_shape(mussel_results$eez_ranked) +
  tm_polygons("suitable_area_km2",
              palette = "-mako",
              title = "Suitable Area (km²)", boarder.col = "black",
              border.lwd = 0.5) +
  tm_layout(main.title ="Suitable Mussel Aquaculuture Area by EEZ",
            main.title.size =1.2,
            legend.outsides =TRUE,
            frame =FALSE) +
  tm_text("rgn", # Label by region
          size = .8, # Adjust size
          col = "white", # Adjust text color
          fontface = "bold", # Labels are bolded
          xmod = -.5) + # Adjust .5 from the left
  tm_compass(type = "4star", 
           position = c("right", "top")) +
tm_scale_bar(position = c("left", "bottom")) 
```

## Discussion

This analysis demonstrates how geospatial suitability modeling can inform strategic aquaculture planning. Central California consistently emerged as the most promising region for both oysters and datemussels.

However, several considerations remain:

  - **Biological constraints:** Growth rates, disease sensitivity, and larval dispersal were not modeled.

  - **Socio-political constraints:** Tribal waters, marine protected areas, and fisheries conflicts are not accounted for.

  - **Water quality:** Nutrient loading, pollution, and harmful algal blooms vary spatially and should be included in future analyses.

Future work could incorporate ecological risk assessments or multi-criteria decision analysis to better integrate environmental and human dimensions.

### References

Hall, S. J., Delaporte, A., Phillips, M. J., Beveridge, M. & O’Keefe, M. Blue Frontiers: Managing the Environmental Costs of Aquaculture [(The WorldFish Center, Penang, Malaysia, 2011)](https://worldfishcenter.org/publication/blue-frontiers-managing-environmental-costs-aquaculture-0).

Maselko, J., Bishop, G., & Murphy, P. (2013). Ghost Fishing in the Southeast Alaska Commercial Dungeness Crab Fishery. North American Journal of Fisheries Management, 33(2), 422–431.doi.org/10.1080/02755947.2013.763875

Gentry, R. R., Froehlich, H. E., Grimm, D., Kareiva, P., Parke, M., Rust, M., Gaines, S. D., & Halpern, B. S. Mapping the global potential for marine aquaculture. Nature Ecology & Evolution, 1, 1317-1324 [(2017)](doi.org/10.1038/s41559-017-0257-9v).︎

GEBCO Compilation Group (2022) GEBCO_2022 Grid [(doi:10.5285/e0f0bb80-ab44-2739-e053-6c86abc0289c)](https://www.gebco.net/data-products/gridded-bathymetry-data/gebco-2022)

Commercially Important Species Occurring in United States (contiguous states). (2025). [(Sealifebase.ca.)] (https://www.sealifebase.ca/country/CountryChecklist.php?c_code=840&vhabitat=commercial)

Oliver, R. (2025, November 25). Homework Assignment 4. [(Github.io.)](https://eds-223-geospatial.github.io/assignments/HW4.html)
